---
title: "Assign CpGs to Intevals"
author: "ks"
date: "2024-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(GenomicFeatures)
```

## TASK: Assign sample DNAm values to genomic intervals. 
### Subset TSSs to chromosomes 1-22 (>95\%)
For each CpG, save interval name, interval start site, and strand. Later we'll summarize the average DNAm by position in the interval and strand.
  

1. Load the saved TSSs, restrict to chromosomes 1-22, and extend interval to 500 bases up- and downstream.
```{r tss-data}
load("data/min_tss_gr.rda")
desired_chromosomes <- paste0("chr",c(as.character(1:22)))
min_tss_gr <- min_tss_gr[seqnames(min_tss_gr) %in% desired_chromosomes]
min_tss_gr <- keepSeqlevels(min_tss_gr, seqlevelsInUse(min_tss_gr), pruning.mode = "coarse")
seqlevels(min_tss_gr)

min_tss_gr <- sort(min_tss_gr)
# Resize to extend 500 bases upstream and downstream (i.e., total width of 1001 bp)
tss_resized <- resize(min_tss_gr, width = 3001, fix = "center")
```

There are `r length(tss_resized)` TSSs.


2. Now grab a DNA methylation dataset. The file below
is setup to read sample IN (aka N2) and save all CpGs that have coverge of 5 or more reads.
```{r dnam-data, message=FALSE}
rmarkdown::render("read-dnam-bedfiles.Rmd", quiet = TRUE)
```

We're starting with `r length(betan2)` measured CpGs.
  
3. Overlap the 2 GRange objects, CpGs with TSS intervals. 

One thing I discovered is that I have overlapping TSS intervals.  I will keep their full overlapping lengths and allow some CpGs to map to multiple intervals.

```{r overlap}
fO <- findOverlaps(betan2,tss_resized)
fO
```
Number of CpGs in intervals: `r length(fO)`  
Number of unique CpGs in intervals: `r length(unique(queryHits(fO)))`   

How are these distributed across different TSS intervals?

```{r describe-fO}
num.in.intervals <- table(table(queryHits(fO)))
num.in.intervals
```

`r num.in.intervals[1]` are in 1 TSS interval  
`r num.in.intervals[2]` are in 2 TSS intervals  
`r num.in.intervals[3]` are in 3 TSS intervals. 
 ...  
 
If I sum these, I get back to the number of unique CpGs.
```{r}
sum(num.in.intervals)
```


Let's get back to assigning CpGs to intervals.
First, I give my CpGs a unique cpg_id, assigning them in 
sequence order. Then, I save the CpGs from my overlap object (queryHits(fO)), and add annotation information from the intervals to which the CpGs mapped (subjectHits(fO)). I will add TSS gene id, TSS location, strand, and indicator whether TSS is in a CpG island.

```{r filt-cpgs}
values(betan2)$cpg_id <- c(1:length(betan2))
beta.by.region <- betan2[queryHits(fO)]
values(beta.by.region)$tssgene_id <- 
      values(tss_resized[subjectHits(fO)])$gene_id

# go back to tss position. Row order for min_tss_gr is identical to row order for tss_resized; tss is one base, start gets it 
    values(beta.by.region)$tss <- 
            start(min_tss_gr[subjectHits(fO)])
    values(beta.by.region)$tsstrand <-
            strand(tss_resized[subjectHits(fO)])
     values(beta.by.region)$TSSinCGI <- 
            values(min_tss_gr[subjectHits(fO)])$TSSinCGI
     

beta.by.region <- keepSeqlevels(beta.by.region, seqlevelsInUse(beta.by.region), pruning.mode = "coarse")
seqlevels(beta.by.region)
```

Let's check the object.
I should have the same numbers of CpGs assigned to
multiple intervals as computed from the above query.

```{r}
table(table(values(beta.by.region)$cpg_id))
```

```{r}
beta.by.region
```


```{r save}
save(beta.by.region,file="data/beta.by.region.rda")
```

Done!

```{r sI}
#sessionInfo()
```

