---
title: "Summarize DNAm and Plot"
author: "ks"
date: "2024-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(GenomicRanges)
library(data.table)
library(ggplot2)
```

# Compute Average DNAm by site across intervals


```{r load-data}
load("data/beta.by.region.rda")

Dt <- data.table(cpgpos = start(beta.by.region),
                    as.data.frame(values(beta.by.region)))
Dt$pos <- ifelse(values(beta.by.region)$tsstrand=="+",
                 start(beta.by.region) - 
                values(beta.by.region)$tss,
                values(beta.by.region)$tss -
                  start(beta.by.region))

head(Dt)
```

Add categorical region average beta variable to stratify plots of average DNA methylation by this categorial variable (avg_beta < 0.05, between 0.05-0.95,  > 0.95). Also stratify by proportion of CpGs in CGI. 
```{r merge-w-region-variables}
load("data/filt_sum.tssregion.rda")

mDt <- merge(Dt,
             filt_sum.tssregion,
             by = "tssgene_id",
             all.y = TRUE)
mDt
```


Let's count the number of CpGs in each position, by strand.
```{r}
nobs<- table(mDt$pos,
             mDt$tsstrand,mDt$cat_region_avg_beta,mDt$region_inCGI)
head(nobs)
```


How many intervals are on the plus strand and how many are on the minus strand, grouped by average beta value of interval?
```{r Dt-counts}
mDt <- mDt[order(,region_inCGI,cat_region_avg_beta,tsstrand)]
mDt[, .(unique_tss_count = uniqueN(tssgene_id)), 
              by = c("region_inCGI",
                     "cat_region_avg_beta","tsstrand")]
```

```{r avgbeta}
avgbeta <- mDt[, .(avg_beta = mean(beta)), 
                  by = c("region_inCGI",
                         "cat_region_avg_beta","tsstrand","pos")]
avgbeta <- avgbeta[order(region_inCGI,cat_region_avg_beta,tsstrand,pos)]
avgbeta
```

```{r}
ggplot(avgbeta) +
        geom_point(aes(x=pos,y=avg_beta,color=factor(region_inCGI), alpha=0.3)) +
        geom_smooth(aes(x=pos,y=avg_beta,color=factor(region_inCGI))) +
        facet_grid(as.factor(cat_region_avg_beta) ~ 
                     as.factor(region_inCGI),
                   scales = "free_y")
```

```{r sI}
sessionInfo()
```

