---
title: "DNA methylation profiles"
author: "ks"
date: "2024-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(GenomicRanges)
library(data.table)
library(ggplot2)
```

# Create Spaghetti plots of DNAm over intervals

```{r load-data}
load("data/beta.by.region.rda")
seqlevels(beta.by.region)
```

Check that I dropped X and Y chromosomes and compute position of CpG relative to TSS in a strand aware fashion.
```{r data-table}
Dt <- data.table(cpgpos = start(beta.by.region),
                           data.frame(values(beta.by.region)))
Dt$pos <- ifelse(values(beta.by.region)$tsstrand=="-",
                  values(beta.by.region)$tss -
                  start(beta.by.region),
                 start(beta.by.region) - 
                 values(beta.by.region)$tss)
head(Dt)
Dt$tsstrand <- factor(Dt$tsstrand)
Dt$TSSinCGI <- factor(Dt$TSSinCGI)
```

Add gene expression deviance for stratifying DNA methylation data. 
```{r merge-w-genedev}
load("data/scdev.rda")

scdev <- data.table(scdev)
scdev <- scdev[!is.na(gene_ids)]
mDt <- merge(x = Dt,
             y = scdev,
             by.x = c("tssgene_id"),
             by.y = c("gene_ids"))
mDt
```

Add interval average beta for stratifying DNA methylation data. 
```{r merge-w-region-variables}
load("data/filt_sum.tssregion.rda")

mDt <- merge(x = mDt,
             y = filt_sum.tssregion)
mDt
```

## Look at DNA methylation profiles of inducible genes 
Plot separately for genes with TSS in CGI and not in CGI

```{r}
depE <- mDt[depmap_esstl=="essential"]
dim(depE)
```

How many genes are there, and what fraction have a TSS in a CpG island
```{r}

```


```{r}
ggplot(depE) +
        #geom_point(aes(x=pos,y=avg_beta,color=genesymbol, group=genesymbol), alpha=0.3) +
        geom_line(aes(x=pos,y=beta,color=genesymbol, group=genesymbol),
                  show.legend = FALSE) +
        facet_grid( ~ TSSinCGI) +
        theme_minimal()
```


```{r}
depE[, uniqueN(genesymbol)]
```

```{r}
first_occurrences <- unique(depE, by = "genesymbol")
table(first_occurrences$TSSinCG)
mean(first_occurrences$TSSinCG)
```

Let's plot 5 genes from each category TSS in/not in CpG island. Visual inspection shows me that the first 10 genes will actually give me 6 with TSS in CpG island, 4 not. Let's start there.
```{r}
depEmatch <- depE[is.element(genesymbol,first_occurrences$genesymbol[1:10])]
```

```{r}
ggplot(depEmatch) +
        geom_point(aes(x=pos,y=beta,color=genesymbol, group=genesymbol), alpha=0.3) +
        geom_line(aes(x=pos,y=beta,color=genesymbol, group=genesymbol),
                  show.legend = FALSE) +
        facet_grid( ~ TSSinCGI)
```


```{r}
ggplot(depEmatch) +
        geom_point(aes(x=pos,y=beta,color=genesymbol, group=genesymbol), alpha=0.3) +
        geom_smooth(aes(x=pos,y=beta,color=genesymbol, group=genesymbol),
                  show.legend = FALSE) +
        facet_grid( ~ TSSinCGI)
```

## Fit smoothed lines for each TSS
Then average fitted values and compare this to the earlier figure





```{r sI}
sessionInfo()
```

