---
title: "Compute PWD for 2 samples"
author: "ks"
date: "2024-10-04"
output: html_document
---

```{r setup-nest, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-nest}
library(GenomicFeatures)
library(AnnotationDbi)
library(rtracklayer)
library(data.table)
```

Read in the beta values from two samples and compute the pairwise differences.
```{r bed_file_dirs}
readdir <- c("~kims/Google Drive/My Drive/Data/hpc-archive/wg_bed/wg_bed5963")
#readdir <- c("/Volumes/extreme ssd/Data/hpc-archive/wg_bed/wg_bed5963")
```

These files are not really BED files because of the header line. Read them in using data.table and create GRanges object.
```{r read_betas}
library(data.table)
f1in <- fread(file.path(readdir,"wg_s1c.bed"))
f2in <- fread(file.path(readdir,"wg_s4c.bed"))
```
`r nrow(f1in)` CpGs are covered in sample 1 and
`r nrow(f2in)` in sample 2.


Make this a GRange object, and restrict to chromosomes 1-22, X, Y, on the primary assembly.
```{r beta_gr}
beta.sample1 <- GRanges(seqnames = f1in$V1,
                   IRanges(start = f1in$V2,width=1))
values(beta.sample1) <- DataFrame(s1.beta = f1in$V4,
                            s1.depth = f1in$V5,
                                s1.m = f1in$V6,
                                s1.u = f1in$V7)

desired_chromosomes <- paste0("chr",c(as.character(1:22), "X", "Y"))
beta.sample1 <- beta.sample1[seqnames(beta.sample1) %in% desired_chromosomes]
# Reset seqinfo to only include chromosomes present in the GRanges object
beta.sample1 <- keepSeqlevels(beta.sample1, seqlevelsInUse(beta.sample1), pruning.mode = "coarse")
nr1 <- length(beta.sample1)

beta.sample2 <- GRanges(seqnames = f2in$V1,
                   IRanges(start = f2in$V2,width=1))
values(beta.sample2) <- DataFrame(s2.beta = f2in$V4,
                            s2.depth = f2in$V5,
                                s2.m = f2in$V6,
                                s2.u = f2in$V7)

beta.sample2 <- beta.sample2[seqnames(beta.sample2) %in% desired_chromosomes]
# Reset seqinfo to only include chromosomes present in the GRanges object
beta.sample2 <- keepSeqlevels(beta.sample2, seqlevelsInUse(beta.sample2), pruning.mode = "coarse")
nr2 <- length(beta.sample2)
```

Sample 1: `r nr1` (`r round(nr1/nrow(f1in),3)*100`\%) CpGs are on the primary assembly  
Sample 2: `r nr2` (`r round(nr2/nrow(f2in),3)*100`\%) CpGs are on the primary assembly

Now require minimum depth of 5 in both samples and compute the pairwise difference.
```{r read-depth-filter}
beta.sample1 <- beta.sample1[values(beta.sample1)$s1.depth > 4]
beta.sample2 <- beta.sample2[values(beta.sample2)$s2.depth > 4]

fO <- findOverlaps(beta.sample1,beta.sample2)
fO

# restrict to overlapping CpGs
overlap.sample1 <- beta.sample1[queryHits(fO)]
overlap.sample2 <- beta.sample2[subjectHits(fO)]
pwd_gr <- overlap.sample1
values(pwd_gr) <- data.frame(values(overlap.sample1),
                             values(overlap.sample2))
rm(overlap.sample1,overlap.sample2)

values(pwd_gr) <- data.frame( pwd = abs(values(pwd_gr)$s1.beta - values(pwd_gr)$s2.beta),
                              values(pwd_gr))
values(pwd_gr)
```




```{r sI-nest}
sessionInfo()
```

