---
title: "Read DNAm .BED files"
author: "ks"
date: "2024-09-27"
output: html_document
---

```{r setup-nest, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-nest}
library(GenomicFeatures)
library(AnnotationDbi)
library(rtracklayer)
#BiocManager::install("TxDb.Hsapiens.UCSC.hg38.knownGene")
library(data.table)
```

Read in the beta values from sample IN.
```{r bed_file_dirs}
readdir <- c("~kims/Google Drive/My Drive/Data/hpc-archive/wg_bed/wg_bed5963")
#readdir <- c("/Volumes/extreme ssd/Data/hpc-archive/wg_bed/wg_bed5963")
```

These files are not really BED files because of the header line. Read them in using data.table and create GRanges object.
```{r beta-n2}
library(data.table)

fin <- fread(file.path(readdir,"wg_s4c.bed"))
```
`r nrow(fin)` CpGs are covered in sample IN (aka N2).


Make this a GRange object, and only save sites with coverage 5 or more reads.
```{r betan2_gr}
betan2 <- GRanges(seqnames = fin$V1,
                   IRanges(start = fin$V2,width=1))
values(betan2) <- DataFrame(beta = fin$V4,
                            depth = fin$V5,
                                m = fin$V6,
                                u = fin$V7)
betan2 <- betan2[values(betan2)$depth > 4]
```
`r length(betan2)` CpGs (`r round(length(betan2)/nrow(fin),3)*100` percent) are covered by 5 or more reads.

Wow. That's still a lot of CpGs!


```{r sI-nest}
sessionInfo()
```

