---
title: "DNAm summaries"
author: "ks"
date: "2024-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(GenomicFeatures)
library(AnnotationDbi)
library(rtracklayer)
#BiocManager::install("TxDb.Hsapiens.UCSC.hg38.knownGene")
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

Get transcription start sites:
From ChatGPT
```{r tss}
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

transcripts_gr <- transcripts(txdb)
tss_gr <- resize(transcripts_gr, width = 1, fix = "start")
tss_gr[1]
start(tss_gr[1])


# Map transcript IDs to gene IDs
gene_tx_map <- AnnotationDbi::select(txdb, keys = as.character(transcripts_gr$tx_id), 
                                     columns = "GENEID", keytype = "TXID")
# Add gene IDs to the TSS GRanges object
tss_gr$gene_id <- gene_tx_map$GENEID[match(as.character(transcripts_gr$tx_id),
                                           gene_tx_map$TXID)]

#Drop rows with missing geneids
tss_gr <- tss_gr[!is.na(tss_gr$gene_id)]
length(tss_gr)
```

These are for all transcripts. I want to reduce them to genes.

First, restrict to chromosomes 1-22, X, Y.  I only want to select TSSs that are
on the primary assembly.
```{r remove_alt_chr}
desired_chromosomes <- paste0("chr",c(as.character(1:22), "X", "Y"))
# Assuming your GRanges object is called granges_obj
tss_gr <- tss_gr[seqnames(tss_gr) %in% desired_chromosomes]
length(tss_gr)
```
Now we'll reduce these to genes.

```{r pick-gene-tss}
# Split the TSS by gene
tss_by_gene <- split(tss_gr, tss_gr$gene_id)

# Find the correct TSS considering strand
min_tss_by_gene <- lapply(tss_by_gene, function(gr) {
                    
                      positive_strand <- gr[strand(gr) == "+"]
                      negative_strand <- gr[strand(gr) == "-"]
  
        # Get minimum for positive strand and maximum for negative strand
        pos_tss <- if (length(positive_strand) > 0)   positive_strand[which.min(start(positive_strand))] else NULL
        neg_tss <- if (length(negative_strand) > 0) negative_strand[which.max(start(negative_strand))] else NULL
  
        # Return Grange for object that isn't NULL
        if (!is.null(pos_tss)) return(pos_tss)
        if (!is.null(neg_tss)) return(neg_tss) 
        })

# Use unlist to combine the GRanges elements into a single GRanges object
min_tss_gr <- unlist(GRangesList(min_tss_by_gene))

# View the result
min_tss_gr
```

Is it correct to assume each gene is only on 1 strand? (I made that assumption)
If I'm wrong, then I've saved the positive strand only.



Now let's grab 500 bases upstream and downstream:
```{r}
# Resize to extend 500 bases upstream and downstream (i.e., total width of 1001 bp)
tss_resized <- resize(tss_gr, width = 1001, fix = "center")
```

Let's overlap sample JN(N1) DNAm values with the TSS regions and summarize.

First read in the beta values from sample JN.
```{r read-data}
readdir <- c("~kims/Google Drive/My Drive/Data/hpc-archive/wg_bed/wg_bed5963")
betajn <- import(file.path(readdir,"wg_s1c.bed"), format = "BED")
```


```{r sI}
sessionInfo()
```

